---
title: "CABANAnet Workshop - Ecological Analysis"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---

# Alpha and Beta Diversity Analysis


````{r message=FALSE, warning=FALSE}
# Load/install required packages
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(phyloseq)

````

```{r}

### Load required data ###

otu_table = read.csv("merged_abundance.csv", 
                          header = TRUE, row.names = 1)

tax_table = read.csv("all_MAGs_taxonomy.csv", 
                     header = TRUE, 
                     row.names = 1
)

sample_data = read.csv("consolidate_metadata.csv", 
                                header = TRUE,
                                row.names = 1,
                                sep = ";"
)

```

### Create Phyloseq Object
```{r}

### transform data to required format ###

# Transposing abundance table, fixing row names, and converting to a matrix
otu_table <- t(otu_table)
rownames(otu_table) <- gsub("^X", "", rownames(otu_table))
otu_table <- as.matrix(otu_table)

# Converting taxonomy table and metadata into a matrix
tax_table <- as.matrix(tax_table)
sample_data<- as.data.frame.matrix(sample_data) 


### Convert data into a phyloseq object ###

# Creating phyloseq components
otu_phyloseq <- otu_table(round(as.matrix(otu_table)), taxa_are_rows = TRUE)
tax_phyloseq <- tax_table(tax_table)
sample_data_phyloseq <- sample_data(sample_data)

# Creating the phyloseq object
ps_data = phyloseq(otu_phyloseq, tax_phyloseq, sample_data_phyloseq)

```

## Alpha Diversity 

### Compute Shannon Index and Observed Richness

```{r}

# Compute diversity measures
diversity_values <- estimate_richness(ps_data, measures = c("Shannon", "Observed"))

# Fix rownames using sample names from ps_data
rownames(diversity_values) <- sample_names(ps_data)


### Prepare data for plot ###

# Merge with metadata
metadata_alpha <- merge(diversity_values,
                      sample_data(ps_data),
                      by = "row.names",
                      sort = FALSE)

# Clean up
rownames(metadata_alpha) <- metadata_alpha$Row.names
metadata_alpha <- metadata_alpha[, -1]


```

### Plot Alpha Diversity

```{r}

# Plot sampling location
sample_location_plot <- ggplot(metadata_alpha, aes(x = sampling_location, y = Shannon, fill = pais)) +
  geom_violin(width= 0.5, alpha = 0.7 ) +  # Violin plot
  geom_boxplot(width = 0.1) +  # Boxplot 
  scale_fill_brewer(palette = "Accent") +
  geom_jitter(width = 0.1, alpha = 0.6) +  # Individual points
  labs(
    x = "Sampling Location",  # Customize x-axis label
    y = "Shannon Index",       # Customize y-axis label
    fill = "Country"  # Custom legend title
  ) +
  theme_bw()

sample_location_plot


# Plot BMI category
bmi_plot <- ggplot(metadata_alpha, aes(x = BMI_Class, y = Shannon, fill = sampling_location)) +
  geom_violin(
    position = position_dodge(width = 0.8), 
    width = 0.7, 
    alpha = 0.7
  ) +
  geom_boxplot(
    width = 0.15, 
    outlier.shape = NA, 
    position = position_dodge(width = 0.8)
  ) +
  geom_jitter(
    position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.8), 
    alpha = 0.6
  ) +
  scale_fill_brewer(palette = "Accent") +
    labs(
    x = "BMI Category",
    y = "Shannon Index",
    fill = "Sampling Location"
  ) +
  theme_bw()

bmi_plot

```


## Beta Diversity 

### Compute Bray-Curtis Distances and PCoa Analysis

```{r}

# Calculate Bray-Curtis distance matrix
bray_dist <- distance(ps_data, method = "bray")

# Perform Principal Coordinates Analysis (PCoA) on the Bray-Curtis distance
pcoa <- ordinate(ps_data, method = "PCoA", distance = bray_dist)


### Prepare data for plot ###

## Eigenvalues ##

# Extract eigenvalues
eigenvalues <- pcoa$values$Eigenvalues

# Calculate the proportion of variance explained by PC1 and PC2
total_variance <- sum(eigenvalues)
pc1_variance <- eigenvalues[1] / total_variance * 100  # PC1 variance explained
pc2_variance <- eigenvalues[2] / total_variance * 100  # PC2 variance explained


## PCoA coordinates ##

# Extract PCoA coordinates
pcoa_scores <- as.data.frame(pcoa$vectors)

# Extract only Axis.1 and Axis.2 from the PCoA coordinates
pcoa_axes <- pcoa_scores[, c("Axis.1", "Axis.2")]

# Rename the columns to PCoA1 and PCoA2
colnames(pcoa_axes) <- c("PCoA1", "PCoA2")


## Merge with metadata ##

metadata_pcoa <- merge(pcoa_axes,
                      sample_data(ps_data),
                      by = "row.names",
                      sort = FALSE)

# Clean up
rownames(metadata_pcoa) <- metadata_pcoa$Row.names
metadata_pcoa <- metadata_pcoa[, -1]


```

### Plot Beta Diversity

```{r}

ggplot(metadata_pcoa, aes(x = PCoA1, y = PCoA2)) +
  geom_point(aes(color = sampling_location, shape = BMI_Class )) + 
 scale_color_brewer(palette = "Accent")+
  theme_bw() + 
    labs(title = paste("PCoA Plot (Bray-Curtis Distances)"),
       x = paste("PC1 (", round(pc1_variance, 1), "%)", sep = ""),
       y = paste("PC2 (", round(pc2_variance, 1), "%)", sep = ""),
       color = "Sampling Location",     # Custom legend title for color
       shape = "BMI Category"  # Custom legend title for shape
       ) 



```

# Compositional Plot with Phyloseq

```{r}

# Your custom color palette
colors <- c("#1E3C2DFF","#78873CFF","#C3C378FF","#695A69FF", 
            "#A58778FF","#B4785AFF","#875A5AFF","#784B69FF", 
            "#695A5AFF", "#877878FF", "#A5A5A5FF", "#CC96B1FF",
            "#9C9EDEFF","#5A5A69FF", "#4B4B69FF")

# Calculate relative abundance
ps_rel_abund <- transform_sample_counts(ps_data, function(x) x / sum(x))

# Aggregate to desired taxonomic level (e.g. Phylum)
ps_rel_phylum <- tax_glom(ps_rel_abund, taxrank = "Phylum")

# Plot 
plot_bar(ps_rel_phylum, fill = "Phylum") +
  scale_fill_manual(values = colors) +  
  facet_wrap(~sampling_location, scales = "free_x") +
  theme_bw() +
  labs(title = "Compositional Barplot at Phylum Level") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```


# Compositional Plot (manual)

### Prepare data for plot
```{r}

# Calculate relative abundance
ps_rel_abund <- transform_sample_counts(ps_data, function(x) x / sum(x))

# Aggregate at the Genus level
ps_genus <- tax_glom(ps_rel_abund, taxrank = "Genus")

# Melt phyloseq object to long format
abundance_df <- psmelt(ps_genus)

# Summarize to find top 15 most abundant taxa
top_genus <- abundance_df %>%
  group_by(Genus) %>%
  summarize(total_abundance = sum(Abundance)) %>%
  arrange(desc(total_abundance)) %>%
  slice_head(n = 15) %>%  
  pull(Genus)

# Reclassify low-abundance taxa as "Other"
abundance_df <- abundance_df %>%
  mutate(Genus = ifelse(Genus %in% top_genus, Genus, "Other"))

# Order taxa by total abundance, with "Other" placed last
genus_levels <- abundance_df %>%
  group_by(Genus) %>%
  summarize(total_abundance = sum(Abundance)) %>%
  arrange(desc(total_abundance)) %>%
  pull(Genus)

genus_levels <- c(setdiff(genus_levels, "Other"), "Other")  # Ensure "Other" is last

abundance_df$Genus <- factor(abundance_df$Genus, levels = genus_levels)

## color palette ##

#create color palette
colors <- c("#1E3C2DFF","#78873CFF","#C3C378FF","#695A69FF", "#A58778FF","#B4785AFF","#875A5AFF","#784B69FF", "#695A5AFF", "#877878FF", "#A5A5A5FF", "#CC96B1FF","#9C9EDEFF","#5A5A69FF", "#4B4B69FF")

# Ensure we have enough colors for the top genera
taxa_colors <- colors[1:(length(genus_levels) - 1)]  # Exclude "Other"

# Assign gray to "Other" and map colors correctly
color_mapping <- c(setNames(taxa_colors, genus_levels[genus_levels != "Other"]), "Other" = "gray80")


```


### Plot Relative Abundance of top 15 most abundant genera
```{r}

ggplot(abundance_df, aes(x = Sample, y = Abundance, fill = Genus)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ sampling_location, scales = "free_x") +  # Locations now appear in the desired order
  scale_fill_manual(values = color_mapping)+
  labs(title = "Compositional Barplot at Genus Level", x = "", y = "Relative Abundance", fill = "Genus") +
  theme_bw() +
  labs(title = "Compositional Barplot at Phylum Level") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))



```


